PHP is a general purpose programming language that is frequently used to run the ``server-side'' code for websites. It was originally created as a simple ``templating language'', based loosely on Perl, to allow outputting HTML with repeated elements. Over the years it has evolved into a fully object-oriented programming language.
\\

PHP should look quite familiar if you've done JavaScript as they are both ``C-based'' languages, meaning that they share a syntax style: semi-colons, curly braces, and brackets.
\\

For now we're just going to run PHP in the command-line like we did in Week 3 with JavaScript. PHP doesn't have a REPL built-in like Node, so we always need to run a file:

\begin{minted}{bash}
    php file.php # run the given file
\end{minted}


\section{Hello, World!}

As is tradition, we should write a ``Hello, World!'' program before moving forward:

\inputminted{php}{01/figures/01/01-hello.php}

As you can see, it's not that different from JavaScript. The main difference is that on line 1 we have to tell PHP that it's PHP. If we don't do this then PHP has a bit of an identity crisis and doesn't know what's going on. This is for historical reasons from when PHP was used as a templating language and was mixed in with HTML.
\\

We ``echo'' the string ``Hello, World!'', this is PHP's equivalent of using \texttt{console.log()} - although it's only useful when used with strings. \texttt{echo} is \textit{not} a function (or method), but instead a \textbf{keyword}. All that really means is that you don't \textit{call} it.
\\

In JavaScript you could get away without semi-colons at the end of lines. PHP isn't nearly so forgiving: if you forget a semi-colon you will get a syntax error and the code will refuse to run.


\section{\texttt{var\_dump}}

As mentioned above \texttt{echo} isn't that useful for values that aren't strings, so we'll be using \texttt{var\_dump} instead. This outputs not only the value we're interested but also relevant information about it.
\\

For example, running the following:

\begin{minted}{php}
    var_dump("Hello");
    var_dump(12 + 12);
    var_dump(true);
\end{minted}

Would give us:

\begin{minted}{text}
    string(5) "Hello"
    int(24)
    bool(true)
\end{minted}

\texttt{var\_dump} \textit{is} a function, so we call it and pass it values.
\\

\texttt{var\_dump} isn't brilliant for working with complex values, so we'll switch to using a nicer way of doing it once we've learned up using PHP libraries.


\section{Variables}

In PHP we don't \textit{declare} variables, we just start using them. This is possible because variables have to start with a \texttt{\$} (this is because of PHP's Perl influence).
\\

This is makes it much easier to accidentally change the values of existing variables, as there is no difference between creating a new variable and reassigning an existing one. So be very careful when naming things.

\phpinputminted{01/figures/01/02-variables}

Variables must start with a dollar, followed by a letter or underscore, followed by any number of letters, numbers, or underscores. Variables names are case-sensitive.
\\

As with JavaScript it is a standard convention to use \texttt{camelCase} for variable names in PHP, although you may see \texttt{snake\_case} used in older code.



\begin{infobox}{Documentation: A Warning}
    Be careful using the official PHP documentation: \textit{anyone} can submit ``User Contributed Notes'' and they often contain code samples that are to be avoided. This is often because the notes were added years ago when PHP was a very different language.
    \\

    As a general rule, don't look at the ``User Contributed Notes'' section: use Stack Overflow if the documentation hasn't cleared it up for you.
\end{infobox}



\section{Additional Resources}

\begin{itemize}[leftmargin=*]
    \item \href{https://www.php.net/manual/en/function.var-dump.php}{PHP: \texttt{var\_dump}}
    \item \href{http://php.net/manual/en/language.variables.basics.php}{PHP: Variable Basics}
    \item \href{https://en.wikipedia.org/wiki/PHP}{Wikipedia: PHP}
    \item \href{https://en.wikipedia.org/wiki/Perl}{Wikipedia: Perl}
\end{itemize}
